---
title: "2018 FRC Match Analysis"
author: Greg Marra
output: github_document
---

A notebook to get FRC Match data from The Blue Alliance API. 

```{r setup_and_get_data}
library(tidyverse)
library(jsonlite)

tba_auth_key <- fromJSON(read_file("tba_auth_key.json"))

# Get Events
url <- "https://www.thebluealliance.com/api/v3/events/2018"
req <- httr::GET(url, httr::add_headers("X-TBA-Auth-Key" = tba_auth_key$tba_auth_key))
json <- httr::content(req, as = "text")
events <- fromJSON(json)

# Get Matches
comp_levels <- c("qm", "qf", "sf", "f")

event_code_list <- events$event_code
output <- vector("list", length(event_code_list))
for (i in seq_along(event_code_list)) {
  url <- paste(c("https://www.thebluealliance.com/api/v3/event/2018",
                 event_code_list[[i]],
                 "/matches"),
               collapse = "")
  req <- httr::GET(url, httr::add_headers("X-TBA-Auth-Key" = tba_auth_key$tba_auth_key))
  json <- httr::content(req, as = "text")
  these_matches <- 0
  tryCatch(these_matches <- fromJSON(json))
  if(length(these_matches) > 0) {
    output[[i]] <- these_matches %>% flatten(recursive = TRUE)
  }
}

matches_flat <- bind_rows(output)
```

# Distribution Of Match Scores

The 2018 FRC game is somewhat unique, in that there is a maximum number of ownership points that can exist in a match, and teams need to tip the accumulation of those points towards their alliance. Early scoring is worth much more in that it can accumulate longer, while a box scored in a Switch or Scale during the last second of the match can only be worth a single point.

Here we look at the results from `r nrow(matches_flat)` matches played so far this year.

```{r match_score_distribution}
matches_flat %>%
  gather(alliances.red.score, alliances.blue.score, key = "alliance", value = "score") %>%
  filter(score >= 0) %>%
  ggplot(aes(score, ..density.., color = parse_factor(comp_level, levels = comp_levels, ordered = TRUE))) +
  geom_density() +
  labs(
    title = "Score distribution across 2018 FRC matches",
    subtitle = "Scores are higher in elims, and closely matched in finals",
    x = "Score",
    y = "Density",
    color = "Comp Level"
  )
```

```{r win_margins}
matches_flat %>%
  filter(alliances.red.score >= 0) %>%
  filter(alliances.blue.score >= 0)%>%
  mutate(
    win_margin = abs(alliances.red.score - alliances.blue.score)
  ) %>%
  ggplot(aes(win_margin, ..density.., color = parse_factor(comp_level, levels = comp_levels, ordered = TRUE))) +
  geom_density() +
  labs(
    title = "Win margins across 2018 FRC matches",
    subtitle = "Matches are closer in finals, as expected",
    x = "Win Margin",
    y = "Density",
    color = "Comp Level"
  )
```

## Ownership Deltas

How big are the swings between ownership?

```{r ownership_deltas}
matches_flat_ownership <- matches_flat %>%
  filter(alliances.red.score >= 0) %>%
  filter(alliances.blue.score >= 0) %>%
  filter(winning_alliance != "") %>%
  mutate(
    red_switch_ownership_pts = score_breakdown.red.autoSwitchOwnershipSec * 2 + score_breakdown.red.teleopSwitchOwnershipSec,
    blue_switch_ownership_pts = score_breakdown.blue.autoSwitchOwnershipSec * 2 + score_breakdown.blue.teleopSwitchOwnershipSec,
    red_scale_ownership_pts = score_breakdown.red.autoScaleOwnershipSec * 2 + score_breakdown.red.teleopScaleOwnershipSec,
    blue_scale_ownership_pts = score_breakdown.blue.autoScaleOwnershipSec * 2 + score_breakdown.blue.teleopScaleOwnershipSec,
    switch_ownership_delta = red_switch_ownership_pts - blue_switch_ownership_pts,
    scale_ownership_delta = red_scale_ownership_pts - blue_scale_ownership_pts)

matches_flat_ownership %>%
  ggplot(aes(switch_ownership_delta, 
             scale_ownership_delta, 
             color = winning_alliance,
             shape = parse_factor(comp_level, levels = comp_levels, ordered = TRUE)),
         alpha = 0.8) +
  geom_point() + 
  geom_abline(slope = -1) +
  labs(
    title = "Ownership Points",
    x = "Switch Ownership Delta (Red - Blue)",
    y = "Scale Ownership Detla (Red - Blue)",
    color = "Winner",
    shape = "Comp Level"
  ) + 
  scale_colour_manual(values = c(red = "red", blue = "blue", NULL = "black"))

matches_flat_ownership %>%
  gather(switch_ownership_delta, scale_ownership_delta, key = "ownership_object", value = "ownership_delta") %>%
  mutate(ownership_object = factor(ownership_object)) %>%
  mutate(ownership_object = fct_recode(ownership_object,
    "Switch" = "switch_ownership_delta",
    "Scale"  = "scale_ownership_delta"
  )) %>%
  ggplot(aes(abs(ownership_delta), color = ownership_object)) +
  geom_freqpoly(binwidth = 6) +
  labs(
    title = "Ownership Points",
    x = "Ownership Delta",
    y = "Match Count",
    color = "Scoring Object"
  )
```


